A Stochastic Quasi-Newton Optimizer for TensorFlow

### 2018-05-06 paper/main.tex
Introduction draft

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{optidef}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{algpseudocode}
\usepackage{algorithm}

@@

\title{A Stochastic Quasi-Newton Optimizer for TensorFlow}
\author{
  Jason Chen \\
  Department of Computer Science \\
  Stony Brook University \\
  Stony Brook, NY 11790 \\
  \texttt{hungrchen@cs.stonybrook.edu} \\
  \And
  David Kraemer \\
  Department of Applied Mathematics\\
  Stony Brook University \\
  Stony Brook, NY 11790 \\
  \texttt{davidkraemer@stonybrook.edu}
}

@@

\begin{enumerate}
  \item Introduce problem that SdLBFGS solves.
  \item TensorFlow current state of the art.
  \item Problems with the TensorFlow state.
  \item Overview of paper.
\end{enumerate}

@@

For a convex function, $\nabla^2 f(x)$ is alwayas positive definite, but this
need not be the case in for nonconvex functions. The work in Wang et al.
\cite{sdlbfgs} shows how the BFGS approximate Hessian may be tweaked to ensure
that positive definiteness is preserved even in nonconvex cases. The overall
stochastic quasi-Newton method is described in Algorithm \ref{alg:sqn}, while
the detailed implementation of SdLBFGS is given in Algorithm \ref{alg:sdlbfgs}.

@@

\begin{algorithm}
  \begin{algorithmic}[1]
    \Function{SdLBFGS}{$g_{k-1}, \ip{s_{j}}_{j=k-p}^{k-2},
    \ip{\bar{y}_{j}}_{j=k-p}^{k-2}, \ip{\rho_{j}}_{j=k-p}^{k-2}$}
    \State{$\textrm{mem} \leftarrow \textit{min}(p, k-1)$}
    \State{$s_{k-1} \leftarrow x_k - x_{k-1}$}
    \State{$y_{k-1} \leftarrow \frac{1}{m_{k-1}}\sum_{i=1}^{m_{k-1}}[g(x_k,
    \xi_{k-1,i}) - g(x_{k-1}, \xi_{k-1,i})]$}
    \State{$\gamma_k \leftarrow \textit{max}\left( \frac{y_{k-1}\trans
    y_{k-1}}{s\trans_{k-1}y_{k-1}}, \delta \right) \geq \delta$}
    \State{%
      $\theta \leftarrow
      \textit{max}\left( 
        \frac{3}{4} \frac{\gamma_{k}\inv s_{k-1}\trans
        s_{k-1}}{\gamma_k\inv s_{k-1}\trans s_{k-1} - s_{k-1}\trans y_{k-1}}, 1
      \right)
    $}
    \Comment{$\theta$ preserves positive definiteness.}
    \State{$\bar{y}_{k-1} \leftarrow \theta y_{k-1} + (1 -\theta \gamma_{k}\inv s_{k-1})$}
    \State{$\rho_{k-1} \leftarrow (s_{k-1}\trans \bar{y}_{k-1})\inv$}
    \For{$i =0 , \dots, \textrm{mem}-1$}
    \State{$\mu_i \leftarrow \rho_{k-i-1} u_i\trans s_{k-i-1}$} 
    \State{$u_{i+1} \leftarrow u_i - \mu_i \bar{y}_{k-i-1}$}
    \EndFor
    \State{$v_0 \leftarrow \gamma_k\inv u_{\textrm{mem}}$}
    \For{$i =0 , \dots, \textrm{mem}-1$}
    \State{$\nu_i \leftarrow \rho_{k-\textrm{mem}+i} v_i\trans
    \bar{y}_{k-\textrm{mem}+i}$}
    \State{$v_{i+1} \leftarrow v_i + (\mu_{\textrm{mem}-i-1} - \nu_i)s_{k-\textrm{mem}+i} $}
    \EndFor
    \State{\Return{$v_p, s_{k-1}, \bar{y}_{k-1}, \rho_{k-1}$}}
    \EndFunction
  \end{algorithmic}
  \caption{%
    The SdLBFGS update step. The resulting output is $H_k g_k = v_p$,
    where $H_k$ is the approximation of the $k$th iterate Hessian and $g_k$ is the
    approximation of the $k$th iterate gradient. Note that the computation of $H_k$
    is implicit, preventing additional storage requirements.
  }
  \label{alg:sdlbfgs}
\end{algorithm}

>>>>

\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{optidef}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{algpseudocode}
\usepackage{algorithm}

@@

\title{A Stochastic Quasi-Newton Optimizer for TensorFlow}

\author{
  Jason Chen \\
  Department of Computer Science \\
  Stony Brook University \\
  Stony Brook, NY 11790 \\
  \texttt{hungrchen@cs.stonybrook.edu} \\
  \And
  David Kraemer \\
  Department of Applied Mathematics\\
  Stony Brook University \\
  Stony Brook, NY 11790 \\
  \texttt{davidkraemer@stonybrook.edu}
}

@@

%\begin{enumerate}
%  \item Introduce problem that SdLBFGS solves.
%  \item TensorFlow current state of the art.
%  \item Problems with the TensorFlow state.
%  \item Overview of paper.
%\end{enumerate}

Limited memory variations of classical second-order minimization algorithms can
prove highly attractive in applications with high-dimensional problems or when
the cost of each iteration step is prohibitive under these classical
formulations.  Specifically, the canonical second-order technique, Newton's
method, involves explicit computation of the Hessian matrix of the objective
function, whose time complexity grows with the square of the parameter
dimension. Even with BFGS, which seeks to reduce the direct computational
difficulty of exactly constructing the Hessian itself, encounters similar
difficulties of scale. The limited memory approach is to bound this overall cost
and ensure that approximate Hessians can be computed on existing hardware.

We implement a variant of stochastic second-order algorithm as described by
Wang, Ma, Goldfarb, and Liu \cite{sdlbfgs} for application in the
\texttt{TensorFlow} library \cite{tensorflow}.  \texttt{TensorFlow} is a
framework for graph-based computation models. Its primary use within the
statistical learning community is in neural networks, which can be readily
implemented by the library. After the model is constructed, the actual training
is performed by an \texttt{Optimizer} object. In this setting, the
\texttt{Optimizer} is presented with a partially observable objective function
and gradient from which to perform optimization. Some typical \texttt{Optimizer}
subclasses which are available in the Python API include
an \texttt{GradientDescentOptimizer}\footnote{%
  See
  \href{https://github.com/tensorflow/tensorflow/blob/master/tensorflow/python/training/gradient_descent.py}{source.}
}% 
and an \texttt{AdagradOptimizer}\footnote{%
  See \href{https://github.com/tensorflow/tensorflow/blob/master/tensorflow/python/training/adagrad.py}{source.}
}%

The presently available optimization algorithms implemented in the Python
library are either first or zeroeth-order methods, requiring access to either
the objective function or its gradient. The conspicuous absence of second-order
methods, such as Newton or quasi-Newton algorithms, has been a subject of much
discussion inside the \texttt{TensorFlow} community\footnote{%
  The primary discussion has occurred surrounding
  \href{https://github.com/tensorflow/tensorflow/issues/446}{Issue \#446}, in
  which contributors have noted that the current infrastructure for developing
  second-order methods is lacking and any implementation would require
  considerable re-inventing the wheel.
.}

Following the suggestions of the community discussion, we present an
implementation of Stochastic damped limited memory BFGS (SdLBFGS) for
\texttt{TensorFlow} through the use of an \texttt{ExternalOptimizerInterface}.
Presently used to wrap existing minimization routines implemented by outside
libraries such as \texttt{SciPy}, it serves as a simple way to engage with the
\texttt{TensorFlow} internals while avoiding a deep dive into its architecture.

Our work proceeds as follows. Section 2 reviews the problem proposed in Wang et
al and gives an explicit formulation of SdLBFGS according to their paper. We
identify small modifications made to the core pseudocode and explain their
necessity. Section 3 describes our SdLBFGS implementation and how it interfaces
with \texttt{TensorFlow}. Section 4 shows preliminary experimental results on
the correctness of our implementation on simple problems. Importantly, we report
that our present implementation is incapable of giving meaningful progress on
training practical models on common datasets. In Section 5, we discuss the
implications of this negative result on future development and specify areas in
which our project may be continued and improved.

Source code for this project can be found at
\url{https://github.com/DavidNKraemer/cse592-project}. Documentation, cruft
removal, and user interface improvements are currently evolving.

@@

For a convex function, $\nabla^2 f(x)$ is always positive semidefinite (see Boyd
and Vandenberghe \S 3.1.4 \cite{boyd}), but this need not be the case in for
nonconvex functions. The work in Wang et al.  \cite{sdlbfgs} shows how the BFGS
approximate Hessian may be tweaked to ensure that positive semidefiniteness is
preserved even in nonconvex cases. The overall stochastic quasi-Newton method is
described in Algorithm \ref{alg:sqn}, while the detailed implementation of
SdLBFGS is given in Algorithm \ref{alg:sdlbfgs}.

@@

\begin{algorithm}
  \begin{algorithmic}[1]
    \Function{SdLBFGS}{$g_{k-1}, \ip{s_{j}}_{j=k-p}^{k-2},
    \ip{\bar{y}_{j}}_{j=k-p}^{k-2}, \ip{\rho_{j}}_{j=k-p}^{k-2}$}
    \State{$\textrm{mem} \leftarrow \textit{min}(p, k-1)$}
    \State{$s_{k-1} \leftarrow x_k - x_{k-1}$}
    \State{$y_{k-1} \leftarrow \frac{1}{m_{k-1}}\sum_{i=1}^{m_{k-1}}[g(x_k,
    \xi_{k-1,i}) - g(x_{k-1}, \xi_{k-1,i})]$}
    \State{$\gamma_k \leftarrow \textit{max}\left( \frac{y_{k-1}\trans
    y_{k-1}}{s\trans_{k-1}y_{k-1}}, \delta \right) \geq \delta$}
    \State{%
      $\theta \leftarrow
      \textit{max}\left( 
        \frac{3}{4} \frac{\gamma_{k}\inv s_{k-1}\trans
        s_{k-1}}{\gamma_k\inv s_{k-1}\trans s_{k-1} - s_{k-1}\trans y_{k-1}}, 1
      \right)
    $}
    \Comment{$\theta$ preserves positive semidefiniteness.}
    \State{$\bar{y}_{k-1} \leftarrow \theta y_{k-1} + (1 -\theta \gamma_{k}\inv s_{k-1})$}
    \State{$\rho_{k-1} \leftarrow (s_{k-1}\trans \bar{y}_{k-1})\inv$}
    \For{$i =0 , \dots, \textrm{mem}-1$}
    \State{$\mu_i \leftarrow \rho_{k-i-1} u_i\trans s_{k-i-1}$} 
    \State{$u_{i+1} \leftarrow u_i - \mu_i \bar{y}_{k-i-1}$}
    \EndFor
    \State{$v_0 \leftarrow \gamma_k\inv u_{\textrm{mem}}$}
    \For{$i =0 , \dots, \textrm{mem}-1$}
    \State{$\nu_i \leftarrow \rho_{k-\textrm{mem}+i} v_i\trans
    \bar{y}_{k-\textrm{mem}+i}$}
    \State{$v_{i+1} \leftarrow v_i + (\mu_{\textrm{mem}-i-1} - \nu_i)s_{k-\textrm{mem}+i} $}
    \EndFor
    \State{\Return{$v_p, s_{k-1}, \bar{y}_{k-1}, \rho_{k-1}$}}
    \EndFunction
  \end{algorithmic}
  \caption{%
    The SdLBFGS update step. The resulting output is $H_k g_k = v_p$,
    where $H_k$ is the approximation of the $k$th iterate Hessian and $g_k$ is the
    approximation of the $k$th iterate gradient. Note that the computation of $H_k$
    is implicit, preventing additional storage requirements.
  }
  \label{alg:sdlbfgs}
\end{algorithm}
